Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-05-19 LICS_2025/monoidal_e_graphs/sections/introduction.tex
camera-ready and arxiv script

Although equality saturation is already a state-of-the-art algorithmic optimisation technique, there is an opportunity to better understand its mathematical foundations and the foundations of e-graphs in particular.
We address this by giving a categorical axiomatisation for e-graphs and their rewriting.
Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and morphisms of Cartesian categories enriched over the category of semilattices (\textit{semilattice-enriched}).
% The morphisms of the latter can also be depicted using \emph{string diagrams}~\cite{Selinger_2010,joyal_geometry_1991, mellies_functorial_2006}: a two-dimensional syntax that encodes equivalence classes of morphism terms.
% Enriched categories~\cite{Borceux_1994,Kelly2022BASICCO} generalise plain categories by imposing more structure making the categorical language more expressive~\cite{Borceux_1994,Kelly2022BASICCO} which will be useful for modelling equivalence between terms as encountered in e-graphs.
Our approach is guided by the established methodology of creating a correspondence between string diagrams and graph rewriting, for which there are existing precedents in the literature~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022-3,fscd,ghica-zanassi2023string}.
The key ingredient of our categorical approach is the semilattice-enrichment\footnote{We could alternatively use the language of idempotent commutative semigroups, which form an equivalent category, but stick to the \enquote{semilattice} terminology for brevity.}.
By using the binary operator of a given semilattice we can operate with (non-empty) \textit{sets} of terms (string diagrams, morphisms); the rest of the paper is about working out the mathematical implications of introducing this enrichment to string diagrams and their concrete combinatorial representation.

@@

The other steps, (c) to (e), represent further elaborations of the e-graph, or e-hypergraph, via the application of more rules.
If we were to apply another rule, $x * 1 \to x$, we would identify the e-class (dashed box) containing $a$ with the e-class containing $a * 1$ (the topmost dashed box) which would require a cycle to be introduced: the left argument of node $*$ would be the same e-class that $*$ belongs to.
Then, an optimal term $a$ could be extracted from the e-graph by traversing all the e-classes, where its optimality is justified by its minimal size as a graph.
To support such cycles we need a richer categorical structure for our string diagrams.
This is obtained by combining this work with the results of~\cite{ghica_rewriting_2023}, and for simplicity we will limit ourselves to the case of acyclic e-graphs, although this is not a limitation of the general theory.
% While equipping the string diagrams which such structure is straightforward, it is out of the scope of this work, and thus we limit ourselves to acyclic e-graphs and e-string diagrams for the purposes of presentation.
Note that while the transformations of the e-graph are guided by the algorithm, the transformations of the string diagrammatic version are guided by the equations of the underlying theory.
A detailed step-by-step application of the equations for the rewrite rules from $(a)$ to $(b)$ and $(b)$ to $(c)$ is given in Figures~\ref{fig:e-graph-example-a-b} and~\ref{fig:e-graph-example-b-c}, respectively.
The equations applied are the ones from Figure~\ref{fig:string-equations} along with the naturality of delete and copy.
Note that the size of the eventual string diagram is comparable to the size of the corresponding e-graph.
% Note that in the final e-graph, (e), a cycle is introduced, making it the representation of infinitely many terms: $a, a*1, (a*1)*1, \ldots$. 
% From the saturated graph the optimal term consisting of only the node $a$ can be then extracted, discarding everything else.

@@

\begin{figure}
\[
    \tikzfig{../figures/categorical-semantics/egraph-translation-1}
\]
\captionsetup{skip=0pt}
\caption{E-graph e-class translation to string diagram.}
\label{fig:e-graph-to-string}
\vspace{-8mm}
\end{figure}
% The subfigure on the right is slightly simplified: if we denote the arity of each $c_i$ as $a_{c_{i}}$, then the dashed box on the right should contain $\sum\limits_{i} a_{c_{i}}$ inputs and each component $c_j$ has $\sum\limits_{i \not = j} a_{c_{i}}$ weakening nodes inside.
% Examples of this translation are given in the second row of Figure \ref{fig:e-graph-example}.
% In particular, note that the translation of the \textit{cyclic} e-graph (e) requires the use of a categorical \textit{trace}, generating a cycle. 
% In this paper, we focus on \textit{acyclic} e-graphs and their corresponding categories; but the above example illustrates that the extension to the cyclic case seems to be routine. As it would involve needless additional presentational complexity we leave it as further work.

@@

\begin{definition}[$\catname{SLat}$-functor]
	Let $\mathbb{C}$ and $\mathbb{D}$ be two $\catname{SLat}$-categories.
	An $\catname{SLat}$-functor $F : \mathbb{C} \to \mathbb{D}$ is defined by the following data.
	\begin{enumerate}
	  \item A mapping $F : \mathbb{C} \to \mathbb{D}$
	  \item An object-indexed family of morphisms in $\catname{SLat}$ $F_{A,B} : \mathbb{C}(A,B) \to \mathbb{D}(FA,FB)$
	\end{enumerate}
	such that certain coherence diagrams commute~\cite{Borceux_1994}.
	$F$ being a morphism in $\catname{SLat}$ has a property $F(f + g) = F(f) + F(g)$.
	An $\catname{SLat}$-functor $F$ defines an equivalence between two $\catname{SLat}$-categories if it is \textit{full}, \textit{faithful}, and \textit{essentially surjective}~\cite{Kelly2022BASICCO}.
\end{definition}

>>>>

Although equality saturation is already a state-of-the-art algorithmic optimisation technique, there is an opportunity to better understand its mathematical foundations and the foundations of e-graphs in particular.
We address this by giving a categorical axiomatisation for e-graphs and their rewriting.
Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and morphisms of Cartesian categories enriched over the category of semilattices (\textit{semilattice-enriched}).
% The morphisms of the latter can also be depicted using \emph{string diagrams}~\cite{Selinger_2010,joyal_geometry_1991, mellies_functorial_2006}: a two-dimensional syntax that encodes equivalence classes of morphism terms.
% Enriched categories~\cite{Borceux_1994,Kelly2022BASICCO} generalise plain categories by imposing more structure making the categorical language more expressive~\cite{Borceux_1994,Kelly2022BASICCO} which will be useful for modelling equivalence between terms as encountered in e-graphs.
Our approach is guided by the established methodology of creating a correspondence between string diagrams and graph rewriting, for which there are existing precedents in the literature~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022-3,fscd,ghica-zanassi2023string}.
The key ingredient of our categorical approach is the semilattice-enrichment.
% \footnote{We could alternatively use the language of idempotent commutative semigroups, which form an equivalent category, but stick to the \enquote{semilattice} terminology for brevity.}
By using the binary operator of a given semilattice we can operate with (non-empty) \textit{sets} of terms (string diagrams, morphisms); the rest of the paper is about working out the mathematical implications of introducing this enrichment to string diagrams and their concrete combinatorial representation.

@@

The other steps, (c) to (d), represent further elaborations of the e-graph, or e-hypergraph, via the application of more rules.
If we were to apply another rule, $x * 1 \to x$, we would identify the e-class (dashed box) containing $a$ with the e-class containing $a * 1$ (the topmost dashed box) which would require a cycle to be introduced: the left argument of node $*$ would be the same e-class that $*$ belongs to.
Then, an optimal term $a$ could be extracted from the e-graph by traversing all the e-classes, where its optimality is justified by its minimal size as a graph.
To support such cycles we need a richer categorical structure for our string diagrams.
This is obtained by combining this work with the results of~\cite{ghica_rewriting_2023}, and for simplicity we will limit ourselves to the case of acyclic e-graphs, although this is not a limitation of the general theory.
% While equipping the string diagrams which such structure is straightforward, it is out of the scope of this work, and thus we limit ourselves to acyclic e-graphs and e-string diagrams for the purposes of presentation.
Note that while the transformations of the e-graph are guided by the algorithm, the transformations of the string diagrammatic version are guided by the equations of the underlying theory.
A detailed step-by-step application of the equations for the rewrite rules from $(a)$ to $(b)$ and $(b)$ to $(c)$ is given in Figures~\ref{fig:e-graph-example-a-b} and~\ref{fig:e-graph-example-b-c}, respectively.
The equations applied are the ones from Figure~\ref{fig:string-equations} along with the naturality of delete and copy.
Note that the degree of sharing in the eventual string diagram is comparable to the degree of sharing in the corresponding e-graph.
% Note that in the final e-graph, (e), a cycle is introduced, making it the representation of infinitely many terms: $a, a*1, (a*1)*1, \ldots$. 
% From the saturated graph the optimal term consisting of only the node $a$ can be then extracted, discarding everything else.

@@

\begin{figure}
\[
    \tikzfig{../figures/categorical-semantics/egraph-translation-1}
\]
\captionsetup{skip=0pt}
\caption{E-graph e-class translation to string diagram.}
\label{fig:e-graph-to-string}
\end{figure}
% The subfigure on the right is slightly simplified: if we denote the arity of each $c_i$ as $a_{c_{i}}$, then the dashed box on the right should contain $\sum\limits_{i} a_{c_{i}}$ inputs and each component $c_j$ has $\sum\limits_{i \not = j} a_{c_{i}}$ weakening nodes inside.
% Examples of this translation are given in the second row of Figure \ref{fig:e-graph-example}.
% In particular, note that the translation of the \textit{cyclic} e-graph (e) requires the use of a categorical \textit{trace}, generating a cycle. 
% In this paper, we focus on \textit{acyclic} e-graphs and their corresponding categories; but the above example illustrates that the extension to the cyclic case seems to be routine. As it would involve needless additional presentational complexity we leave it as further work.

@@

\begin{definition}[$\catname{SLat}$-functor]
	Let $\mathbb{C}$ and $\mathbb{D}$ be two $\catname{SLat}$-categories.
	An $\catname{SLat}$-functor $F : \mathbb{C} \to \mathbb{D}$ is defined by the following data.
	\begin{enumerate}
	  \item A mapping $F : \mathbb{C} \to \mathbb{D}$
	  \item An object-indexed family of morphisms in $\catname{SLat}$ $F_{A,B} : \mathbb{C}(A,B) \to \mathbb{D}(FA,FB)$
	\end{enumerate}
	such that certain coherence diagrams commute~\cite{Borceux_1994}.
	$F_{A,B}$ being a morphism in $\catname{SLat}$ has a property $F_{A,B}(f + g) = F_{A,B}(f) + F_{A,B}(g)$.
	An $\catname{SLat}$-functor $F$ defines an equivalence between two $\catname{SLat}$-categories if it is \textit{full}, \textit{faithful}, and \textit{essentially surjective}~\cite{Kelly2022BASICCO}.
\end{definition}
